---
title: "Detecting Execution Mistakes in Weight Lifting Exercises Using On-body sensing and Machine Learning"
author: "Renaud Dufour"
date: "May, 2015"
output: html_document
---

# Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of 
data about personal activity relatively inexpensively. These type of devices are part of the quantified self
movement – a group of enthusiasts who take measurements about themselves regularly to improve their
health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do
is quantify how much of a particular activity they do, but they rarely quantify how well they do it.
In this project, we use data from accelerometers on the belt, forearm, arm, and dumbell of 6
participants. They were asked to perform perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:

* exactly according to the specification (Class A)
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C)
* lowering the dumbbell only halfway (Class D)
* throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.
More information is available from the website [here](http://groupware.les.inf.puc-rio.br/har]) (see the section on the Weight Lifting Exercise Dataset).

Goal of the project is to predict the manner in which people did the exercise ("classe" variable).

# Data

The dataset provided can be obtained from my [Github Repo](https://github.com/duf59/plm_project) in the source_data/ folder.
It is composed of two files : 

* pml-training.csv : a training set containing 19622 observations of 160 variables including the dependent variable "classe".
* pml-testing.csv : the validation set containing 20 observations of 159 variables for which the "classe" is to be predicted.

We first load these two datasets :

```{r load data}
data <- read.csv("./source_data/pml-training.csv", na.string = c(NA,""))
validation  <- read.csv("./source_data/pml-testing.csv", na.string = c(NA,""))
```

# Preprocessing

A summary of the dataset reveals that out of the 159 predictors, 100 present 98% of missing values in the *data* dataset (same predictors present 100% missing values in the *validation* dataset). These predictors are removed :

```{r remove NA}
idx1 <- which(sapply(data,function(x) sum(is.na(x))/nrow(data))>0.95)
data <- data[,-idx1]
```

Among the 59 predictors left, We remove *X* which corresponds to the row number.
Besides, the purpose being to predict the *classe* outcome from the sensors informations,
we do not use predictors related to time and date:

```{r remove time variables}
data <- subset(data, select=-c(X,raw_timestamp_part_1,raw_timestamp_part_2,cvtd_timestamp,new_window,num_window))
```

These preprocessing steps lead to the following 53 predictors (for more details about their meaning refer to the [original website](http://groupware.les.inf.puc-rio.br/har]) and related informations):

```{r display initial predictors}
names(data)
```

We then look at the correlation Matrix. Below we print the number of strong correlations (>0.80) for each predictor (not including autocorrelation):

```{r find correlated variables}
corMatrix <- abs(cor(subset(data, select=-c(user_name,classe))))>.8
diag(corMatrix) <- FALSE # do not take into account autocorrelations
colSums(corMatrix)       # display number of high correlation for each variable
```

For threshold of 0.80, looking more clesely at the correlation matrix reveals the following strong pairwise correlations:

* *total_accel_belt* with *accel_belt_y* and *accel_belt_z* (last 2 also correlated)
* *magnet_belt_x* with *accel_belt_x* and *pitch_belt* (last 2 also correlated)
* *gyro_arm_x* with *gyro_arm_y*
* *magnet_arm_x* with *accel_arm_x*
* *magnet_arm_y* with *magnet_arm_z*
* *pitch_dumbbell* with *accel_dumbbell_x*
* *yaw_dumbdell* with *accel_dumbbell_z*
* *gyros_forearm_z* with *gyros_forearm_y* and *gyros_dumbbell_z* and *gyros_dumbbell_x* (last 2 also correlated)

A vector of variables to remove to reduce pairwise correlation can also be obtained as follows:

```{r flag correlated variables}
library(caret)
flags <- findCorrelation(cor(subset(data, select=-c(user_name,classe))),cutoff = .80)
names(subset(data, select=-c(user_name,classe)))[flags]
```

Which confirms the previous observations.

# Machine learning

Before training a model, we split the data into a training and testing sets.
The training set is to be used for training the model while the testing set will be use to provide an unbiased estimate
of the model accuracy.
Note that we do not use here the validation set of 20 observations because these predictions are to be uploaded on Coursera for project grading.

Splitting is done as folows (ratio 75/25):

```{r split}
set.seed(3764) # for reproducibility
inTrain<- createDataPartition(data$classe, p = 0.75, list = FALSE)
training <- data[inTrain,]
testing  <- data[-inTrain,]
```

## Random forest model using all predictors

We train a random forest on the training set, keeping all 53 predictors.
10-fold cross validation error is used along with different values of the mtry parameter (number of variables allowed for each split):

```{r randomForest fit, eval=FALSE}
# code not evaluated in Markdown
# takes about 40 minutes to run on intel core i5-4200U and 8Gb Ram.
numFolds <- trainControl(method = "cv", number = 10)
rffit    <- train(classe ~ ., method="rf", data=training, trControl = numFolds)
```
```{r load rf model, echo=FALSE}
# read saved random forest model
rffit <- readRDS(file = "rfmodel-no-num_window.rds")
```
```{r print rf model}
# read saved random forest model
rffit
```

The best RF is obtained for mtry=29, leading to a CV accuracy of 99.2%.
Plot below shows the variable importance (top 10 predictors):

```{r varImpr}
library(randomForest)
varImpPlot(rffit$finalModel, n.var = 10, main = "Random Forest Model - top 10 predictors")
```

We then apply the model on the hold out testing set in order to estimate the model accuracy:

```{r randomForest evaluation}
testPred <- predict(rffit, newdata = testing)
confusionMatrix(testPred,testing$classe)
```

The estimated model accuracy on the testing set is 0.993 (error rate 0.007).
This is consistent with the CV accuracy meaning we are not overfitting the training set.

The model was then applied to the 20 observations of the validation set (result not shown here because subject to the Coursera Honor Code.)

## Random forest model using 9 predictors

The variable importance of the previous model, along with the pairwise correlations discused previously, suggest that not all predictors are necessary to perform a proper prediction on this dataset. Therefore, considering correlations and variable importance, the number of predictors was iteratively decreased until the following training set was obtained:

```{r small dataset}
training2 <- subset(training, select=-c(user_name,
                                        gyros_forearm_x,gyros_forearm_y,gyros_forearm_z,
                                        gyros_arm_x,gyros_arm_y,gyros_arm_z,
                                        gyros_dumbbell_x,gyros_dumbbell_y,gyros_dumbbell_z,
                                        gyros_belt_x,gyros_belt_y,gyros_belt_z,
                                        yaw_dumbbell, pitch_dumbbell,
                                        magnet_arm_x, magnet_arm_y, magnet_arm_z,
                                        pitch_arm,
                                        yaw_forearm,
                                        accel_arm_x, accel_arm_y, accel_arm_z,
                                        total_accel_forearm, total_accel_arm, total_accel_belt, total_accel_dumbbell,
                                        accel_belt_x,accel_belt_y,accel_belt_z,
                                        accel_forearm_x,accel_forearm_y,accel_forearm_z,
                                        magnet_forearm_x,magnet_forearm_y,magnet_forearm_z,
                                        accel_dumbbell_x, accel_dumbbell_y, accel_dumbbell_z,
                                        magnet_belt_x, magnet_belt_y, magnet_belt_z,
                                        yaw_arm, roll_arm))

```

Remaining predictors are :

```{r}
names(training2)
```

Then train a random forest using these predictors :

```{r randomForest small dataset, eval=FALSE}
# code not run in Markdown - takes about 15min
numFolds <- trainControl( method = "cv", number = 10)
cpGrid   <- expand.grid( .mtry = seq(3,10,by = 1))
rffit2   <- train(classe ~ ., method="rf", data=training2, trControl = numFolds, tuneGrid = cpGrid)
```
```{r load rf model small dataset, echo=FALSE}
# read saved random forest model
rffit2 <- readRDS("rfmodel-training3.rds")
```
```{r print rf model small dataset}
# read saved random forest model
rffit2
```

This model achieves 0.985 CV accuracy for mtry=3, which is confirmed by predicting on the held out testing set:

```{r randomForest evaluation small dataset}
testPred <- predict(rffit2, newdata = testing)
confusionMatrix(testPred,testing$classe)
```

Note that this model with only 9 predictors exhibit the same accuracy in predicting the 20 observations of the validation set.
